% ============================================================
%  APPENDIX C : Glossary of Blockchain Terms
% ============================================================
\chapter{Glossary of Blockchain Terms}
\label{app:glossary}

\noindent\textbf{Declaration of Contributions:} This appendix was jointly prepared by all team members.

\bigskip

\begin{longtable}{l p{10cm}}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
\endhead
Block & A data structure containing a batch of transactions, a timestamp, and a link (hash) to the previous block. \\
Blockchain & A chain of cryptographically linked blocks forming an immutable, distributed ledger. \\
Consensus & Agreement among network nodes on the valid state of the ledger. \\
Cryptographic Hash & A fixed-size digital fingerprint of arbitrary data; any change to the input changes the output completely. \\
Decentralisation & Distribution of control and data across many nodes with no single authority. \\
Double Spending & The fraudulent act of spending the same digital token twice. \\
Fork & A divergence in the blockchain when two valid blocks are produced simultaneously. \\
Genesis Block & The very first block in a blockchain (Block 0). \\
Immutability & The property that once data is recorded on the blockchain, it cannot be altered or deleted. \\
Merkle Tree & A binary tree of hashes used to efficiently summarise and verify large sets of data. \\
Mining & The process of solving a computational puzzle to add a new block (in PoW systems). \\
Node & A computer participating in the blockchain network. \\
Nonce & A number that miners vary to find a valid block hash in Proof of Work. \\
Smart Contract & A self-executing program on a blockchain that enforces contractual terms automatically. \\
Validator & A node that validates and proposes blocks in Proof of Stake systems. \\
Wallet & Software that manages a user's cryptographic keys and blockchain addresses. \\
\bottomrule
\end{longtable}
